Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-08-09 documents/ms.tex
getting ready to resubmit

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth, clip=true]{figures/koi_lc_demo.pdf}
\caption[A light curve with a GP model.]
{Light curve of KIC \kepexample, an active star with a rotation period of
$\sim$ \kepexampleperiod\ days.
The solid line shows a fit to the data using a Gaussian process model with a QP
covariance kernel function.}
\label{fig:GP_example}
\end{center}
\end{figure}

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_acfprior_03_10.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique with an informed, ACF-based prior.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
Since the posterior PDFs of rotation periods are often non-Gaussian,
    the points plotted here are maximum {\it a-posteriori} results.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
The ACF-informed prior on rotation period used to generate these results is
    described in \textsection \ref{sec:GP_prior}.
    }
\label{fig:compare_mcmc_acfprior}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_noprior_02_13.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique with an uninformative prior.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
Since the posterior PDFs of rotation periods are often non-Gaussian,
    the points plotted here are maximum {\it a-posteriori} results.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
An uninformative prior, flat in the natural log of the rotation period between
    0.5 and 100 days was used to generate these results.
    }
\label{fig:compare_mcmc_noprior}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/corner_plot.pdf}
\caption{Marginal posterior PDFs of the QP GP model parameters, fit to the
    simulated light curve in figure \ref{fig:demo_lc}.
    The solid line in the period panel shows the injected period.
    This figure was made using \texttt{corner.py} \citep{Corner}.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth, clip=true]{figures/demo_ACF.pdf}
\caption[ACF of a simulated light curve.]
{An autocorrelation function of the simulated light curve shown in figure
\ref{fig:demo_lc}.
The vertical solid blue line shows the period measured using the ACF method (20.4
days) and the pink dashed line shows the period that was used to simulate the
light curve (\aigrainexampleperiod\ days).}
\label{fig:demo_acf}
\end{center}
\end{figure}

@@

\response{We first applied a high-pass filter to the light curves, removing
long-term trends that could produce large peaks at long periods in the
periodogram.
We used the 3rd order Butterworth filter in {\tt scipy} with a cut-off
of 35 days, attenuating signals with periods greater than this threshold.
This filter was applied in frequency space, rather than period space, and full
attenuation was attained at zero frequency.
In other words, the level of attenuation increased smoothly with period, for
periods above 35 days but no periods were fully suppressed.
Experimentally, we found the 35 day cut off removed the largest number of long
period outliers while preserving shorter period signals.
We also tried 50, 45, 40 and 30 day cut-offs but the 35 day cut-off value
minimised the RMS and MAD of the LS periodogram results.}
For each simulated light curve, we computed a LS periodogram\footnote{LS
periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.} over a
grid of 10,000 periods, evenly spaced in frequency, between 1 and 100 days.
We adopted the period of the highest peak in the periodogram as the measured
rotation period.
\response{The uncertainties on the rotation periods were calculated using the
following equation for the standard deviation of the frequency
\citep{Horne1986, Kovacs1981}:
\begin{equation}
    \sigma_{\nu} = \frac{3\pi\sigma_N}{2N^{1/2}TA},
\end{equation}
where $A$ is the amplitude of the signal of highest power, $\sigma_N$ is the
variance of the time series, with the signal of highest power removed, $N$
is the number of observations and $T$ is the timespan of the data.
These formal uncertainties are only valid in the case that the noise is white,
the data are evenly sampled and there is only one signal present.
Since there are multiple signals present in these light curves, this formal
uncertainty is an underestimate of the true uncertainty.
}
Figure \ref{fig:pgram_compare} shows the resulting recovered rotation periods
as a function of true period.
\response{The MRD of the periodogram-recovered periods is \percentpgramMAD\%,
slightly worse than the ACF and GP methods (see table \ref{tab:MADs} for
a side-by-side comparison with the other methods).
The filter suppresses signals with periods greater than 35 days but doesn't
eliminate them altogether, thus allowing high amplitude frequencies to be
detectable.
For this reason, some erroneous long periods are recovered by the periodogram
method.
These results still show a marked improvement in comparison to periodograms
calculated from unfiltered data.
Simply adopting the highest peak of a periodogram calculated from the {\it
unfiltered} simulated data resulted in an RMS of \oldpgramRMS\ and relative
MAD of \oldpercentpgramMAD.}
% The LS periodogram method drastically overestimates some rotation periods
% because long-term trends unrelated to rotation still produced excess power
% at long periods, despite our attempts to remove these trends via filtering.
% {\bf Filtering a light curve before attempting to measure a rotation period is
% not always advisable because it may remove real signals from the data.
% The simulated light curves used in this experiment had rotation periods of up
% to 50 days, and thus applying a high pass filter with a 35 day cut off left
% most of these periods unaffected.
% Of course, this would not be the case if there were many slower rotators in
% this sample.
% In some cases a signal greater than 35 days was recovered, sometimes correctly
% and sometimes erroneously.
% }
% Filtering is not analogous to using priors in Bayesian inference as filtering
% completely removes signals; if a periodic signal is filtered out there is no
% hope of recovering it.
% Using priors that favour shorter periods over longer ones still allows for the
% detection of long periods if they are strongly supported by the data.}

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{compare_pgram.pdf}
\includegraphics[width=6in, clip=true]{figures/compare_pgram_filtered_35.pdf}
\caption[LS periodogram results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using a LS periodogram technique.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
In many cases a large peak at a long period was present in the periodogram,
    producing a significant over-estimate of the period.
    }
\label{fig:pgram_compare}
\end{center}
\end{figure*}

@@

\begin{table*}
\begin{center}
    \caption{Rotation periods and \Kepler\ input catalogue (KIC) properties
    for the \Kepler\ objects of interest. The full table is available online.}
\begin{tabular}{lcccc}
    KOI number & Period (days) & KIC T$_{\mathrm{eff}} (K)$ & KIC $\log(g)$ &
    KIC $[$Fe$/$H$]$ \\
    \hline
1 & 25$^{+23}_{-6}$ & $5126^{+160}_{-138}$ & $4.61^{+0.03}_{-0.11}$ & $-0.12\pm0.3$\\
2 & 20$\pm4$ & $5850\pm50$ & $4.46\pm0.03$ & $-0.15\pm0.1$\\
3 & 28$^{+3}_{-0.2}$ & $6350\pm80$ & $4.02\pm0.01$ & $0.26\pm0.08$\\
6 & 23$^{+3}_{-15}$ & $6225^{+114}_{-158}$ & $4.17^{+0.06}_{-0.05}$ & $-0.04^{+0.1}_{-0.2}$\\
7 & 25$\pm2$ & $5543\pm79$ & $4.08\pm0.01$ & $0.44\pm0.1$\\
    ... & ... & ... & ... & ... \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:koi_periods}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_koi_01_23.pdf}
\caption[Comparison with McQuillan results.]
{A comparison of our GP rotation period measurements to those of
\citet{Mcquillan2013}.
The data points are coloured by the range of variability measured by
    \citet{Mcquillan2013}, defined as the interval between the 5th and 95th
    percentiles of normalized flux per period bin in millimagnitudes.}
\label{fig:mcquillan}
\end{center}
\end{figure*}

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth, clip=true]{koi_lc_demo.pdf}
\caption[A light curve with a GP model.]
{Light curve of KIC \kepexample, an active star with a rotation period of
$\sim$ \kepexampleperiod\ days.
The solid line shows a fit to the data using a Gaussian process model with a QP
covariance kernel function.}
\label{fig:GP_example}
\end{center}
\end{figure}

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{comparison_acfprior_03_10.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique with an informed, ACF-based prior.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
Since the posterior PDFs of rotation periods are often non-Gaussian,
    the points plotted here are maximum {\it a-posteriori} results.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
The ACF-informed prior on rotation period used to generate these results is
    described in \textsection \ref{sec:GP_prior}.
    }
\label{fig:compare_mcmc_acfprior}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{comparison_noprior_02_13.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique with an uninformative prior.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
Since the posterior PDFs of rotation periods are often non-Gaussian,
    the points plotted here are maximum {\it a-posteriori} results.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
An uninformative prior, flat in the natural log of the rotation period between
    0.5 and 100 days was used to generate these results.
    }
\label{fig:compare_mcmc_noprior}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{corner_plot.pdf}
\caption{Marginal posterior PDFs of the QP GP model parameters, fit to the
    simulated light curve in figure \ref{fig:demo_lc}.
    The solid line in the period panel shows the injected period.
    This figure was made using \texttt{corner.py} \citep{Corner}.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth, clip=true]{demo_ACF.pdf}
\caption[ACF of a simulated light curve.]
{An autocorrelation function of the simulated light curve shown in figure
\ref{fig:demo_lc}.
The vertical solid blue line shows the period measured using the ACF method (20.4
days) and the pink dashed line shows the period that was used to simulate the
light curve (\aigrainexampleperiod\ days).}
\label{fig:demo_acf}
\end{center}
\end{figure}

@@

\response{We first applied a high-pass filter to the light curves, removing
long-term trends that could produce large peaks at long periods in the
periodogram.
We used the 3rd order Butterworth filter in {\tt scipy} with a cut-off
of 35 days, attenuating signals with periods greater than this threshold.
This filter was applied in frequency space, rather than period space, and full
attenuation was attained at zero frequency.
In other words, the level of attenuation increased smoothly with period, for
periods above 35 days but no periods were fully suppressed.
Experimentally, we found the 35 day cut off removed the largest number of long
period outliers while preserving shorter period signals.
We also tried 50, 45, 40 and 30 day cut-offs but the 35 day cut-off value
minimised the RMS and MAD of the LS periodogram results.}
For each simulated light curve, we computed a LS periodogram\footnote{LS
periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.} over a
grid of 10,000 periods, evenly spaced in frequency, between 1 and 100 days.
We adopted the period of the highest peak in the periodogram as the measured
rotation period.
\response{The uncertainties on the rotation periods were calculated using the
following equation for the standard deviation of the frequency
\citep{Horne1986, Kovacs1981}:
\begin{equation}
    \sigma_{\nu} = \frac{3\pi\sigma_N}{2N^{1/2}TA},
\end{equation}
where $A$ is the amplitude of the signal of highest power, $\sigma_N$ is the
variance of the time series, with the signal of highest power removed, $N$
is the number of observations and $T$ is the timespan of the data.
These formal uncertainties are only valid in the case that the noise is white,
the data are evenly sampled and there is only one signal present.
Since there are multiple signals present in these light curves, this formal
uncertainty is a noisy estimate of the true uncertainty.
}
Figure \ref{fig:pgram_compare} shows the resulting recovered rotation periods
as a function of true period.
\response{The MRD of the periodogram-recovered periods is \percentpgramMAD\%,
slightly worse than the ACF and GP methods (see table \ref{tab:MADs} for
a side-by-side comparison with the other methods).
The filter suppresses signals with periods greater than 35 days but doesn't
eliminate them altogether, thus allowing high amplitude frequencies to be
detectable.
For this reason, some erroneous long periods are recovered by the periodogram
method.
These results still show a marked improvement in comparison to periodograms
calculated from unfiltered data.
Simply adopting the highest peak of a periodogram calculated from the {\it
unfiltered} simulated data resulted in an RMS of \oldpgramRMS\ and relative
MAD of \oldpercentpgramMAD.}
% The LS periodogram method drastically overestimates some rotation periods
% because long-term trends unrelated to rotation still produced excess power
% at long periods, despite our attempts to remove these trends via filtering.
% {\bf Filtering a light curve before attempting to measure a rotation period is
% not always advisable because it may remove real signals from the data.
% The simulated light curves used in this experiment had rotation periods of up
% to 50 days, and thus applying a high pass filter with a 35 day cut off left
% most of these periods unaffected.
% Of course, this would not be the case if there were many slower rotators in
% this sample.
% In some cases a signal greater than 35 days was recovered, sometimes correctly
% and sometimes erroneously.
% }
% Filtering is not analogous to using priors in Bayesian inference as filtering
% completely removes signals; if a periodic signal is filtered out there is no
% hope of recovering it.
% Using priors that favour shorter periods over longer ones still allows for the
% detection of long periods if they are strongly supported by the data.}

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{compare_pgram.pdf}
\includegraphics[width=6in, clip=true]{compare_pgram_filtered_35.pdf}
\caption[LS periodogram results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using a LS periodogram technique.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
In many cases a large peak at a long period was present in the periodogram,
    producing a significant over-estimate of the period.
    }
\label{fig:pgram_compare}
\end{center}
\end{figure*}

@@

\begin{table*}
\begin{center}
    \caption{Rotation periods and \Kepler\ input catalogue (KIC) properties
    for \nkoi\ \Kepler\ objects of interest. The full table is available online.}
\begin{tabular}{lcccc}
    KOI number & Period & KIC T$_{\mathrm{eff}} $ & KIC $\log(g)$ &
    KIC $[$Fe$/$H$]$ \\
    & (days) & (K) & dex & dex \\
    \hline
1 & 25$^{+23}_{-6}$ & $5126^{+160}_{-138}$ & $4.61^{+0.03}_{-0.11}$ & $-0.12\pm0.3$\\
2 & 20$\pm4$ & $5850\pm50$ & $4.46\pm0.03$ & $-0.15\pm0.1$\\
3 & 28$^{+3}_{-0.2}$ & $6350\pm80$ & $4.02\pm0.01$ & $0.26\pm0.08$\\
6 & 23$^{+3}_{-15}$ & $6225^{+114}_{-158}$ & $4.17^{+0.06}_{-0.05}$ & $-0.04^{+0.1}_{-0.2}$\\
7 & 25$\pm2$ & $5543\pm79$ & $4.08\pm0.01$ & $0.44\pm0.1$\\
    ... & ... & ... & ... & ... \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:koi_periods}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{comparison_koi_01_23.pdf}
\caption[Comparison with McQuillan results.]
{A comparison of our GP rotation period measurements to those of
\citet{Mcquillan2013}.
The data points are coloured by the range of variability measured by
    \citet{Mcquillan2013}, defined as the interval between the 5th and 95th
    percentiles of normalized flux per period bin in millimagnitudes.}
\label{fig:mcquillan}
\end{center}
\end{figure*}
